---
title: "SOC-GA 2332 Intro to Stats Lab 10 Solution"
author: "Di Zhou"
date: "4/9/2021"
output:
  html_document:
    df_print: paged
    theme: paper
    highlight: textmate
    toc: true
  pdf_document: 
    toc: true
---


<style type="text/css">

body{ 

    font-size: 16px;
    line-height: 1.7em;
    <!-- text-align: justify; -->

}

blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 16px;
    border: solid 1px;
}

h1 { font-size: 32px; }

h2 { font-size: 24px; }

h3 { font-size: 20px; }

</style>

<br>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(pacman)
p_load(tidyverse, stargazer, kableExtra, gridExtra, effects, gmodels, MASS, nnet)

# Load data
load("data/support_level_df.RData")

```

## Part 2: Bi-variate Associations (Contingency Tables)  

```{r, echo=FALSE}

# Create variables for contingency tables
support_df <- support_df %>%
  mutate(# Covert dummies to categorical variables
         gender = ifelse(female == 0, "male", "female"),
         race = ifelse(black == 1, "black", "white"))

```

### Part 2 Exercise

Recall that the $\chi^2$ statistic is defined as: 

$$\chi^2 = \sum\frac{(f^o - f^e)^2}{f^e},$$
where $f^o$ is the observed frequency and $f^e$ is the expected frequency.  

You are given the following contingency table of support levels and gender: 

```
   Cell Contents
|-------------------------|
|                       N |
|              Expected N |
|           N / Row Total |
|           N / Col Total |
|         N / Table Total |
|-------------------------|
 
Total Observations in Table:  1000 

                         | support_df$gender 
support_df$support_level |    female |      male | Row Total | 
-------------------------|-----------|-----------|-----------|
                       1 |       105 |       147 |       252 | 
                         |   123.228 |   128.772 |           | 
                         |     0.417 |     0.583 |     0.252 | 
                         |     0.215 |     0.288 |           | 
                         |     0.105 |     0.147 |           | 
-------------------------|-----------|-----------|-----------|
                       2 |       109 |       126 |       235 | 
                         |   114.915 |   120.085 |           | 
                         |     0.464 |     0.536 |     0.235 | 
                         |     0.223 |     0.247 |           | 
                         |     0.109 |     0.126 |           | 
-------------------------|-----------|-----------|-----------|
                       3 |       275 |       238 |       513 | 
                         |   250.857 |   262.143 |           | 
                         |     0.536 |     0.464 |     0.513 | 
                         |     0.562 |     0.466 |           | 
                         |     0.275 |     0.238 |           | 
-------------------------|-----------|-----------|-----------|
            Column Total |       489 |       511 |      1000 | 
                         |     0.489 |     0.511 |           | 
-------------------------|-----------|-----------|-----------|

```
1. State you null and alternative hypotheses of the $\chi^2$ test;  
2. Calculate the $\chi^2$ statistic using the formula above;
3. Calculate the p-value of your test statistic. *Hint*: (a) recall that the degree of freedom is calculated by $\text{df} =(\text{nrow}−1)·(\text{ncol}−1)$, (b) search `pchisq`   

**Answer**  
  
  1. $H_0$: Support level and gender are independent; $H_a$: Support level and gender are not independent.  
```{r }
# 2. Calculate the Chi-squared statistic by hand:
(105 - 123.228)^2/123.228 + (147-128.772)^2/128.772 + (109-114.915)^2/114.915 + (126-120.085)^2/120.085 + (275-250.857)^2/250.857 + (238-262.143)^2/262.143
# chi-sq = 10.41945

# 3. p-value
pchisq(10.41945, 2, lower.tail = F)
# OR: 1 - pchisq(10.41945, 2, lower.tail = T)

```

You can use the `CrossTable` function in the `gmodels` package to obtain the above table with expected frequencies and chi-square test results:

```{r }
CrossTable(support_df$support_level, support_df$gender, prop.chisq = F, expected = T)

```

---

## Part 3: Ordered Logit Regression Model  


### Part 3.2 Exercise  

Interpret the exponentiated regression coefficients of `eduy` and `age` of `ologit4`. 

```
##      eduy       age    female     black 
## 2.3726365 0.7965623 2.8403564 0.9533769
```

**Answer:**

Holding others as constant, for one unit increase in years of education, the odds of one moving towards the next higher support level will increase by a factor of 2.37, which means the odds will increase by 137%. This effect is statistically significant at the p<0.01 level.

Holding others as constant, for one unit increase in one's age, the odds of one moving towards the next higher support level will "increase" by a factor of 0.80, which means the odds will decrease by 20%. This effect is statistically significant at the p<0.01 level.

---

## Part 4: Multinomial Logit Regression Model

### 4.2 Coefficients Interpretation

### Part 4.2 Exercise  

Interpret the coefficients of `eduy` and `age` from the above output of exponentiated regression coefficients of `mlogit4`.

```
##   (Intercept)     eduy       age   female     black
## 2   1.9535790 1.820853 0.8464970 2.060538 0.9811196
## 3   0.3756124 3.983080 0.6977276 5.400878 0.9056941
```

**Answer:**

Holding others at constant, for one unit increase in years of education, the relative risk of being neutral to same-sex marriage instead of opposing it will increase by a factor of 1.82, which means the relative risk will increase by 82%. This effect is statistically significant at the p<0.01 level.

Holding others at constant, for one unit increase in years of education, the relative risk of being supportive to same-sex marriage instead of opposing it will increase by a factor of 3.98, which means the relative risk will increase by 298%. This effect is statistically significant at the p<0.01 level.

Holding others at constant, for one unit increase in age, the relative risk of being neutral to same-sex marriage instead of opposing it will "increase" by a factor of 0.85, which means the relative risk will decrease by 15%. This effect is statistically significant at the p<0.01 level.

Holding others at constant, for one unit increase in age, the relative risk of being supportive to same-sex marriage instead of opposing it will "increase" by a factor of 0.70, which means the relative risk will decrease by 30%. This effect is statistically significant at the p<0.01 level.

---  