---
title: "SOC-GA 2332 Intro to Stats Lab 3 Exercise solution"
author: "Di Zhou"
date: "2/19/2021"
output:
  html_document:
    df_print: paged
    theme: paper
    highlight: textmate
    toc: true
  pdf_document: 
    toc: false
---

<style type="text/css">

body{ 

    font-size: 16px;
    line-height: 1.7em;
    <!-- text-align: justify; -->

}

blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 16px;
    border: solid 1px;
}

h1 { font-size: 32px; }

h2 { font-size: 24px; }

h3 { font-size: 20px; }

</style>

<br>

```{r setup, include = T, message = F, warning = F}
knitr::opts_chunk$set(echo = TRUE)

# Load package
library(tidyverse)

# Import data
weight_df <- read.csv("data/weight.csv")

```

### Part 1 Exercise  

Create a sampling distribution of the sample mean  
  (1) Create a virtual population that follows a normal distribution of mean = 0 and variance = 1, with 10,000 observations. *Hint*: use `rnorm()`.  
  (2) Use this population and a `for`-loop, get a sampling distribution of the sample mean by sampling 100 observations from the population for 1,000 times.  
  (3) Plot a histogram of your sampling distribution of the sample mean. You can add lines for mean and SD as demonstrated in the figure above.  

*Note:* Use the `set.seed()` function every time you perform a random process. That is, add the `set.seed()` function before every part of your code that involves a random process and run the `set.seed()` function together with that code for every run.     
    
```{r part1-exercise}

set.seed(202102)

# generate population df
pop_norm <- rnorm(n = 10000, mean = 0, sd = 1) %>% as_tibble()

# create a "container" object to save the mean
mean_vec <- vector(mode = "numeric", length = 1000)

# repeatly sample from the pop df and save the mean 
for (i in 1:1000){

  sample <- pop_norm %>% sample_n(size = 100, replace = FALSE)

  mean_vec[i] <- mean(sample$value)
  
}



# Mean and SD of the sampling distribution
mean = mean(mean_vec)
sd_pos = mean + sd(mean_vec)
sd_neg = mean - sd(mean_vec)


# Plot
mean_vec %>% 
  as_tibble() %>%
  ggplot(aes(value)) +
  geom_histogram(binwidth = 0.01, fill = "grey", color = "black") +
  geom_vline(aes(xintercept = sd_neg), color = "red") +
  geom_vline(aes(xintercept = mean), color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = sd_pos), color = "red") +
  labs(title = "Sampling Distribution of the Sample Mean (1000 iterations of sample n = 100)",
       subtitle = "Mean marked by the dashed line, 1 standard deviation around the mean marked by solid lines")


```

### Part 2.1 Exercise

With $\mu_0 = 0$, $\overline y = 1.54$, sample $n = 27$, $s = 3.25$, derive:  
  (1) The *t* test statistic  
  (2) The critical value of *t* given $H_0$ is true  
  (3) Your rejection region  
  (4) P-value  
  (5) Your conclusion of the significance test  
  
  
```{r part2.1-exercise}  

# First, calcuate standard error of the mean
se = 3.25/sqrt(27)    # 0.6254628

# Calculate t test statistics given null is true
observed_t = (1.54 - 0)/se  # 2.462177

# Critical value of t given null is true
qt(p = 0.5*0.05, df = 26)   # -2.055529

# Rejection region is: (-inf, -2.055529) U (2.055529, inf)

# P-value
2*(1 - pt(q = observed_t, df = 26))  # 0.02075478

# Conclusion:
# Observed t is in the rejection region, so we reject the null
# Alternatively, our p-value is smaller than alpha level, so we reject the null

# You can also try the example here: https://www.geogebra.org/m/b85v7zww

```


### Calculation and functions in R

```{r calculation and functions}

# generate a random number (n = 30)
set.seed(1001)
sample <- rnorm(n = 30, mean = 2, sd = 1)

# hand code to get the sample mean 
sum(sample)/length(sample)

# should equal to:
mean(sample)

# create a function that calculate sample mean
sample_mean <- function(x){
  mean = sum(x)/length(x)
  return(mean)
}

# try the function
sample_mean(sample)

# hand code to get the sample standard deviation
sample_var <-  sum( ( sample - mean(sample) )^2 ) / (length(sample)-1) 
sample_sd <- sqrt(sample_var)
sample_sd

# Use base r function
sd(sample)

# hand-code function
sd_example <- function(x){
  mean = sum(x)/length(x)
  variance = sum((x-mean)^2)/(length(x)-1)
  standarddeviation = sqrt(variance)
  return (standarddeviation)
}

sd_example(sample)



```   

### Part 2.2 Exercise

The institution that offers therapy programs to the anorexic patients claims that their treatment will lead to a weight increase of 4 lbs. Use the `weight_df` data and with $\alpha = 0.05$, perform both a two-tailed and a one-tailed test:
  $$H_0: \mu_{\text{change}} = 4$$
<p align="center">
against
</p>   
  
  $$H_{\text{a1}}: \mu_{\text{change}} \neq 4 \text{   and    } H_{\text{a2}}: \mu_{\text{change}} < 4 $$  
Report your hypothesis testing result. *Hint:* Make sure you put correct arguments for your `t.test()` function! (Are you testing for "two.sided", "less", or "greater"? What's your `mu`?) 

```{r part2.2-exercise}

# two tailed t-test
t.test(weight_df$change, mu = 4, alternative = "two.sided", conf.level = 0.95)
# P-value is larger than 0.05, we do not reject the null

# one tailed t-test
t.test(weight_df$change, mu = 4, alternative = "less", conf.level = 0.95)
# P-value is larger than 0.05, we do not reject the null

```

### Part 3 Exercise  

Perform a two-sample two-tailed t-test for the difference between therapy b and c:

$$H_0: \mu_b - \mu_c = 0$$ 
<p align="center">
against
</p>  
$$H_{a}:\mu_b - \mu_c \ne 0$$

```{r part3-exercise}

# Filter data for each therapy
weight_b <- weight_df %>% filter(therapy == "b")
weight_c <- weight_df %>% filter(therapy == "c")

# two tailed two indep sample t-test
t.test(
  x = weight_b$change,          # mean value vector from the first sample
  y = weight_c$change,          # mean value vector from the second sample
  mu = 0,                       # mean difference given by your null hypothesis
  alternative = "two.sided"     # direction of alternative hypothesis
)

# P-value is larger than 0.05, we do not reject the null

```

