---
title: "SOC-GA 2332 Intro to Stats Lab 6 Solution"
author: "Di Zhou"
date: "3/12/2021"
output:
  html_document:
    df_print: paged
    theme: paper
    highlight: textmate
    toc: true
  pdf_document: 
    toc: true
---


<style type="text/css">

body{ 

    font-size: 16px;
    line-height: 1.7em;
    <!-- text-align: justify; -->

}

blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 16px;
    border: solid 1px;
}

h1 { font-size: 32px; }

h2 { font-size: 24px; }

h3 { font-size: 20px; }

</style>

<br>

---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)

library(pacman)
p_load(tidyverse, stargazer, kableExtra, coefplot)

```


```{r import and run model, include=F}

# Load cleaned and recoded df
load("data/earnings_df.RData")

# Estimate Nested Models
m1 <- lm(earn ~ age_recode, 
         data = earnings_df)

m2 <- lm(earn ~ age_recode + edu,
         data = earnings_df)

m3 <- lm(earn ~ age_recode + edu + female,
         data = earnings_df)

m4 <- lm(earn ~ age_recode + edu + female + black + other,
         data = earnings_df)

m5 <- lm(earn ~ age_recode + edu + female + black + other + edu*female,
         data = earnings_df)

```
### Part 3 Exercise

Plot the effect of `age` on `earn` according to Model 5. Post your plot on Slack.

```{r exercise-ploteffect }

# first, we create a dataframe with all predictor variables
# with only the key IV varies
pred_age <- tibble(age_recode = seq(20, 65, 3)) %>%         
  mutate(edu = mean(earnings_df$edu, na.rm = T),
         female = mean(earnings_df$female, na.rm = T),       
         black = mean(earnings_df$black, na.rm = T),
         other = mean(earnings_df$other, na.rm = T))

# use `predict` to predict the Y
predicted_earning_byage <- predict(m5,                      # the model you are using
                             pred_age,                # the df you use for predicting
                             interval = "confidence", # set CI
                             level = 0.95)

# bind the columns
pred_age_result <- cbind(pred_age, predicted_earning_byage)

# check df
head(pred_age_result, 10) %>% kbl("html") %>% kable_classic_2(full_width = F)
 
# Plot
pred_age_result %>% 
  ggplot(aes(x = age_recode, y = fit)) +
  geom_line() + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +   # add 95% CI
  labs(x = "Age",
       y = "Predicted Earnings") +
  ggtitle("Predicted Earnings by Age")

```

---

### Part 4 Exercise

1. (You can use pencil and paper) Show that 
$$\frac{(SSE_\text{restricted} - SSE_\text{full})/df_1}{SSE_\text{full}/df_2} = \frac{(R^2_\text{full} - R^2_\text{restricted})/df_1}{(1 - R^2_\text{full})/df_2}$$

  Answer: Since the $df1$ and $df2$ part of the two equations are the same, we only need to show that $$\frac{SSE_\text{restricted} - SSE_\text{full}}{SSE_\text{full}} = \frac{R^2_\text{full} - R^2_\text{restricted}}{1 - R^2_\text{full}}$$
  Given that $R^2 = \frac{TSS - SSE}{TSS} = 1 - \frac{SSE}{TSS}$, the right-hand side of the above equation can be written as:
  
  $$\frac{R^2_\text{full} - R^2_\text{restricted}}{1 - R^2_\text{full}} $$ 
  $$ = \frac{(1 - \frac{SSE_\text{full}}{TSS}) - (1 - \frac{SSE_\text{restricted}}{TSS})}{1 - (1 - \frac{SSE_\text{full}}{TSS})}$$
  $$ =  \frac{\frac{SSE_\text{restricted}}{TSS} - \frac{SSE_\text{full}}{TSS}}{\frac{SSE_\text{full}}{TSS}}$$
  $$= \frac{SSE_\text{restricted} - SSE_\text{full}}{SSE_\text{full}}$$

2. Create a new variable `age_sq` in `earnings_df` that is the square term of `age_recode`. Estimate Model 6: earn ~ age + edu + female + race + edu*female + age_square  

  Then, perform a F-test between `m5` and `m6`. What is your null and alternative hypothesis? What's your decision of the F-test?

```{r f-test}

# create age_recode square term
earnings_df <- earnings_df %>%
  mutate(age_sq = age_recode^2)

# estimate model
m6 <- lm(earn ~ age_recode + edu + female + black + other + edu*female + age_sq,
         data = earnings_df)

# check model
stargazer(m5, m6, type="text", omit.stat=c("ser", "f"))

# F test
anova(m5, m6)


```
  
  $H_0$: the coefficient of the additional term equals zero, or $\beta_\text{age_sq} = 0$.  
  
  $H_a$: the coefficient of the additional term is different from zero, , or $\beta_\text{age_sq} \neq 0$.  
  
  Since the P-value is larger than 0.05, we cannot reject the null that $\beta_\text{age_sq} = 0$. This means the new IV added in `m6` does not increase the predictive power. Thus `m5` is preferred in comparison to `m6`.
  

